NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is the most common negatively
charged lipid in eukaryotic membranes.
PS lipids interact with signaling and other proteins via
electrostatic interactions and direct binding, and induce
membrane fusion and phase separation together with calcium ions.
Molecular details of these phenomena are not well understood,
because accurate models to interpret the experimental data have not
been available. Here, we collect a set of experimental NMR data that
can be used together with molecular dynamics (MD) simulations
to interpret the lipid headgroup structures and details of ion binding
in pure PS, and mixed PS:PC, lipid bilayers. We use the open collaboration
method to collect data from a wide range of available MD models of PS lipids.
None of the models reproduce the NMR data within experimental accuracy,
but the best headgroup models suggest that the carboxyl group in the serine headgroup
does not rotate freely. In line with the previous results for PC lipids,
none of the force fields correctly captures the cation binding affinity.
In contrast to PC lipids, the response of PS headgroups
to bound ions  in the tested MD simulation models
qualitatively differs from experiments. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

>>>>

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with signaling and other proteins via
electrostatic interactions and direct binding, and can induce
membrane fusion and phase separation in the presence of calcium ions.
Molecular details of these phenomena are not well understood,
because accurate models to interpret the experimental data have not
been available. Here, we gather a set of experimental NMR data that 
can, together with molecular dynamics (MD) simulations, be used
to decipher the lipid headgroup structures and ion binding
in pure PS and mixed (here PS:PC) lipid bilayers. Using the open collaboration
method, we extract data from wide range of available PS MD models (force fields) and perform a comparison to NMR results.
%or: Using the open collaboration method, we collect a database of MD simulations using wide range of available PS models (force fields) and perform a comparison to NMR results
We find that none of the models reproduce the NMR data within experimental accuracy,
but the best ones suggest that the carboxyl group in the serine headgroup
does not rotate freely. In line with the previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity.
Furthermore, the response of PS headgroups to bound ions
\emph{qualitatively} differs from experiments. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}
